Inequality in socioemotional skills: a cross-cohort comparison

### 2019-02-28 codes/R/paper/seskills_jpube_rev.Rnw
new version of text and tables - incorporates most reviewer comments

@@

@
\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on our novel scale, separately by cohort. The EFA is based on the decomposition of the polychoric correlation matrix, and uses weighted least squares and oblimin rotation.
\par}
\end{table}

@@

@
\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of Rutter items from the BCS. The EFA is based on the decomposition of the polychoric correlation matrix, and uses weighted least squares and oblimin rotation. Items denoted by * are used in the 11-item scale, and items denoted by ** are combined in the 11-item scale.
\par}
\end{table}

@@

@
\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of SDQ items from the BCS. The EFA is based on the decomposition of the polychoric correlation matrix, and uses weighted least squares and oblimin rotation. Items denoted by * are used in the 11-item scale.
\par}
\end{table}

@@

\begin{table}[ht]
\centering\setstretch{1.2} \footnotesize
\caption{Parameterisations for measurement invariance}\label{tab:invparam}
\begin{tabular}{L{.18\textwidth}L{.45\textwidth}L{.3\textwidth}}
\toprule
\textbf{Invariance level}     & \textbf{Description} & \textbf{Restrictions} \\ \midrule \\
Configural (WE{\textTheta})   
    & \bi[label=$\cdot$] \item Minimally restrictive model for identification \ei
    & For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\
                      \bm{\kappa}=\bm{0} \\
                      \bm{\nu}=\bm{0} \\
                      \text{diag}(\bm{\Psi})=\bm{I} \end{array}$
                      \\ \\
Threshold invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ to be equal across groups
                  \item Statistically equivalent to configural (when measures have 3 categories or less)\ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\ 
                                                \bm{\kappa}=\bm{0} \end{array}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \end{array}$
                   \\ \\
Threshold and Loading invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
                  \item Allows comparison of latent factor variances \ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline For all groups: $\bm{\kappa}=\bm{0}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
                   \\ \\
Threshold, Loading, and Intercept invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
                  \item Restricts intercepts $\nu$ to zero in both groups
                  \item Allows comparison of latent factor variances \emph{and} means \ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline For all groups: $\bm{\nu}=\bm{0}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\kappa}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
\\ \\ \bottomrule
\end{tabular}
\justify {\scriptsize \emph{Notes}: Adapted from \cite{Wu2016a}.
\par}
\end{table}

>>>>

\begin{table}[ht]
\centering\setstretch{1.2} \footnotesize
\caption{Parameterisations for measurement invariance}\label{tab:invparam}
\begin{tabular}{L{.18\textwidth}L{.4\textwidth}L{.45\textwidth}}
\toprule
\textbf{Invariance level}     & \textbf{Description} & \textbf{Restrictions} \\ \midrule \\
Configural (WE{\textTheta})   
    & \bi[label=$\cdot$] \item Minimally restrictive model for identification \ei
    & For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\
                      \bm{\kappa}=\bm{0} \\
                      \bm{\nu}=\bm{0} \\
                      \text{diag}(\bm{\Psi})=\bm{I} \end{array}$
                      \\ \\
Threshold invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ to be equal across groups
                  \item Statistically equivalent to configural (when measures have 3 categories or less)\ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline For all groups: $\begin{array}{|l} \text{diag}(\bm{\Phi}) = \bm{I} \\ 
                                                \bm{\kappa}=\bm{0} \end{array}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \end{array}$
                   \\ \\
Threshold and Loading invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
                  \item Allows comparison of latent factor variances \ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline For all groups: $\bm{\kappa}=\bm{0}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\nu}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
                   \\ \\
Threshold, Loading, and Intercept invariance
    & \bi[label=$\cdot$] \item Restricts thresholds $\tau$ and loadings $\lambda$ to be equal across groups
                  \item Restricts intercepts $\nu$ to zero in both groups
                  \item Allows comparison of latent factor variances \emph{and} means \ei
    & $\tau_{1,ci} = \tau_{1,c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline $\tau_{2,ci} = \tau_{2,c^\prime i}$ for non-binary items, $\forall c, c^\prime$
                      \newline $\lambda_{ci} = \lambda_{c^\prime i}$ for all items, $\forall c, c^\prime$
                      \newline For all groups: $\bm{\nu}=\bm{0}$
                      \newline For ref. group $A$: $\;\begin{array}{|l} \bm{\kappa}_A=\bm{0} \\
                                                  \text{diag}(\bm{\Sigma}_A)=\bm{I} \\
                                                  \text{diag}(\bm{\Phi}_A) = \bm{I} \end{array}$ 
\\ \\ \bottomrule
\end{tabular}
\justify {\scriptsize \emph{Notes}: Adapted from \cite{Wu2016a}.
\par}
\end{table}

@@

@
\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on our novel scale, separately by cohort. The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}). We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.
\par}
\end{table}

@@

@
\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of Rutter items from the BCS. Items denoted by * are used in the 11-item scale. The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}). We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.
\par}
\end{table}

@@

@
\justify {\scriptsize \emph{Notes}: The table displays the factor loadings obtained from exploratory factor analysis (EFA) on the full set of SDQ items from the BCS. Items denoted by * are used in the 11-item scale. The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}). We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.
\par}
\end{table}

@@

### 2019-02-28 codes/R/paper/text/paper_jpube_1introlit.tex
new version of text and tables - incorporates most reviewer comments

The importance of cognition in predicting life course success is well established in the economics literature. However, in recent years the role played by `non-cognitive' traits is being increasingly investigated. These traits include constructs as different as psychological and preference parameters such as social and emotional skills, locus of control and self-esteem, personality traits (e.g. conscientiousness), and risk aversion and time preferences. Given the vastity of this literature, we briefly review below the main papers on the determinants and consequences of socio-emotional traits which are more directly related to our work, and we refer to other sources for more exhaustive reviews \citep{Borghans2008,Almlund2011,Goodman2015,Kautz2014}.

@@

Very few papers attempt to make cross-cohorts comparisons about the importance of socio-emotional skills. \citet{Blanden2007} -- one of the closes study to ours -- examine cognitive skills, non‐cognitive traits, educational attainment and labour market attachment as mediators of the decline in inter-generational mobility in UK between the 1958 and the 1970 cohorts. The authors take great care in selecting non-cognitive items to be as comparable as possible across cohorts, from the Rutter scale at age 10 for the 1970 cohort and from the Bristol Social Adjustment Guide for the 1958 cohort; however, they do not carry out a formal test of measurement invariance and the do not construct factor scores fully comparable across cohorts as we do. Another paper related to ours is the one by \citet{Reardon2016}, who study recent trends in income, racial, and ethnic school gaps in several dimensions of school readiness, including academic achievement, self-control, and externalizing behavior, at
kindergarten entry, using comparable data from the Early Childhood Longitudinal Studies (ECLS-K and ECLS-B) for cohorts born from the
early 1990s to the 2000–2010 period in the US. They find that readiness gaps narrowed modestly from 1998 to 2010, particularly between high- and low-income students and between White and Hispanic students. Lastly, \citet{Deming2017} uses a comparable set of skill measures and covariates across survey waves for the NLSY79 and the NLSY97, and finds that the labour market return to social skills was much greater in the 2000s than in the mid-1980s and 1990s.

@@

In sum, even if the literature on the determinants and consequences of socio-emotional skills has been booming, most papers use skills measured in late childhood or in adolescence; and no paper in economics formally tests for invariance of measurements across different groups and constructs fully comparable scores. In this paper, we use measures of child socio-emotional development at age 5, hence before formal schooling starts; and we construct comparable scales across the two cohorts we study (the 1970 and the 2000 British cohorts), so that we can investigate changes in inequality in early development, their determinants, and consequences, in a parallel fashion.

>>>>

The importance of cognition in predicting life course success is well established in the economics literature. However, in recent years the role played by `non-cognitive' traits is being increasingly investigated. These traits include constructs as different as psychological and preference parameters such as social and emotional skills, locus of control and self-esteem, personality traits (e.g. conscientiousness), and risk aversion and time preferences. Given the vastness of this literature, we briefly review below the main papers on the determinants and consequences of socio-emotional traits which are more directly related to our work, and we refer to other sources for more exhaustive reviews \citep{Borghans2008,Almlund2011,Goodman2015,Kautz2014}.

@@

Very few papers attempt to make cross-cohorts comparisons about the importance of socio-emotional skills. \citet{Blanden2007} -- one of the closest study to ours -- examine cognitive skills, non‐cognitive traits, educational attainment and labour market attachment as mediators of the decline in inter-generational income mobility in UK between the 1958 and the 1970 cohorts. The authors take great care in selecting non-cognitive items to be as comparable as possible across cohorts, from the Rutter scale at age 10 for the 1970 cohort and from the Bristol Social Adjustment Guide for the 1958 cohort; however, they do not carry out a formal test of measurement invariance and the do not construct factor scores fully comparable across cohorts as we do. Another paper related to ours is the one by \citet{Reardon2016}, who study recent trends in income, racial, and ethnic school gaps in several dimensions of school readiness, including academic achievement, self-control, and externalizing behavior, at
kindergarten entry, using comparable data from the Early Childhood Longitudinal Studies (ECLS-K and ECLS-B) for cohorts born from the
early 1990s to the 2000–2010 period in the US. They find that readiness gaps narrowed modestly from 1998 to 2010, particularly between high- and low-income students and between White and Hispanic students. Lastly, \citet{Deming2017} uses a comparable set of skill measures and covariates across survey waves for the NLSY79 and the NLSY97, and finds that the labour market return to social skills was much greater in the 2000s than in the mid-1980s and 1990s.

@@

In sum, even if the literature on the determinants and consequences of socio-emotional skills has been booming, most papers use skills measured in late childhood or in adolescence; and no paper in economics formally tests for invariance of measurements across different groups and constructs fully comparable scores. In this paper, we use measures of child socio-emotional development at age 5, hence at the very beginning of formal schooling; and we construct comparable scales across the two cohorts we study (the 1970 and the 2000 British cohorts), so that we can investigate changes in inequality in early development, their determinants, and consequences, in a parallel fashion.
